Ownership types in theory and practice (in Rust)

### 2022-11-30 paper.tex
More progress

% TODO: wording of value and types used interchangedly here, maybe incorrect
First, one must differentiate between values that are stored on the stack and those stored on the heap.

The primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack since their sizes are small and known at compile time.
Values are stored on the heap if their size is dynamic and therefore not known at compile time.

The different representations of strings in Rust provide a good example.
Rust supports the immutable string literals \emph{\&str}, which are stored on the stack, and the type \emph{String}, which is stored as a vector of bytes, heap allocated, and therefore growable. \cite{rust-by-example}

The \emph{String} type in Rust is a good example of such a type.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
\end{lstlisting}

% This, however, applies only to data types that are stored on the heap.

In the following piece of Rust code, one variable is assigned to another.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
\end{lstlisting}
Since their sizes are known at compile time, they are simply copied into a new variable instead of moved.

@@

>>>>

% TODO: wording of value and types used interchangedly here, maybe incorrect
% First, one must differentiate between values that are stored on the stack and those stored on the heap.
In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
Values are stored on the heap if their size is dynamic and therefore not known at compile time.

% TODO: a good example of what?
The different representations of strings in Rust provide a good example.
Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable. \cite{rust-by-example}

% TODO: footnote that main class etc. have been omitted for sake of simplicity
The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
% TODO: explain what Rust traits are
Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that

However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
% TODO: go into detail on how strings are stored on heap.
The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
Thus, copying the whole string into \emph{y} could be very inefficient.
% TODO: avoid and at beginning of sentence
And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
After the move, \emph{x} is no longer valid.
Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.

% TODO: add section explaing borrowing, since it's related ot move

@@

The following adjustement to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x.clone();
println!("x: {}, y: {}", x, y);
\end{lstlisting}
